### 2017-01-05 Chapters/experiments.tex
Last corrections from hans, also add list of figures and tables

The following pages contain the full annotated sequences of the region downstream \textit{rpsB} after the insertion of \textit{gfp\_mut2-cassette} and \textit{mCherry-cassette}, respectively (third line in Fig.~\ref{fig:Cloning}).
The primers annealing positions are indicated using the same color code as in the description of the primers above.
The primers reported in the publication of Bakshi \textit{et al.}~\cite{bakshi_superresolution_2012} are indicated for information.

@@

After the strains were made using the protocol described in Material and Methods~\ref{sec:methods_strain}, the downstream region of the \textit{rpsB} gene was amplified and sent for sequencing.
Alignments of the results showed they exhibit the expected sequences, except for a couple of single-base mutations.
The raw results of the sequencing are reproduced below.

@@

In Fig.~\ref{fig:suppinfo_caract}, we show growth curves obtained on M9 supplemented with 0.2\% glucose.
As for the construction of Bakshi \textit{et al.}~\cite{bakshi_superresolution_2012}, our strains possess a wild-type growth rate (Fig.~\ref{fig:suppinfo_caract}A), indicating that the gfp-tagging or mCherry-tagging of the ribosomal S2 subunit does not impede the functioning of the ribosome.
It is known that \textit{E.~coli} cultures exhibit a high autofluorescence in the GFP bandwidth~\cite{mihalcescu_green_2015}, but the fluorescence of the \textit{rpsB-gfp} strain is at least 4-time higher than the \textit{wild-type} (Fig.~\ref{fig:suppinfo_caract}B).
Contrary to green autofluorescence, red autofluorescence of \textit{E.~coli} cultures is very low (Fig.~\ref{fig:suppinfo_caract}B) so the corrected signal level for mCherry is even more stronger.
Overall, both strains possess enough fluorescence to allow quantification at the population level, all the more so at the cell level because most part of the autofluorescence is concentrated in the medium outside the cell~\cite{mihalcescu_green_2015}.

In Fig.~\ref{fig:suppinfo_caract}C, we display the ratio between the (corrected) fluorescence of the strain and the absorbance (corrected means that we substracted the autofluorescence of the wild-type strain).
As a first approximation, this ratio can be used as a proxy for the fluorescence concentration in the cells, therefore for the ribosome concentration.
The instability at the beginning of the experiment (before 400~min) is characteristic of the difficulty to obtain robust estimations when the absorbance of the culture is low~\cite{zulkower_robust_2015}.
The RFU/Abs ratio is stable in the interval [400,600]~min for both \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains, and can be interpreted as exponential steady-state growth.
Note that this short time interval (a couple of generations) of readability before approaching the stationary phase is one of the main problem preventing the use of batch conditions to obtain robust data about growth transitions.
Indeed, most of our attempt to perform robust growth transitions in this region were unsuccessful.

Results for \textit{rpsB-gfp} and \textit{rpsB-mCherry} start to differ when glucose is exhausted and the cells enter stationary phase.
As would be expected, the fluorescence of the \textit{rpsB-gfp} strain starts to slowly decrease as GFP proteins are degraded or photobleached.
On the contrary, the fluorescence of the \textit{rpsB-mCherry} strain quickly triples when entering stationary phase, at a rate that is way higher than the physical limit imposed by the protein synthesis rate~\cite{scott_interdependence_2010}.

\begin{figure}[p]
\centering
\includegraphics[scale=0.985]{./Fig/suppinfo_caract}
\caption{\small
\textbf{Growth curves in M9~0.2\% glucose for the \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains.}
Growth and measurements were performed at 37$^\circ$C on a 96-well microplate in a Tecan infinite 200 pro reader.
We monitored 4 wells for each of the WT, \textit{rpsB-gfp}, and \textit{rpsB-mCherry} strains.
Curves were time shifted to correct for variability in the inoculation process (-50 min for WT, and -20 min for \textit{rpsB-mCherry}).
\textit{(A)}~600~nm absorbance as measured by the Tecan microplate reader, corrected for background by removing M9 absorbance.
Curve superposition indicates similar growth rates between the WT and the modified strains.
\textit{(B)}~Fluorescence measured in each well, expressed in Relative Fluorescence Units (RFU).
Green fluorescence (485~nm excitation, 535~nm emission) is measured for the WT (dashed lines) and the \textit{rpsB-gfp} (solid lines) strains.
Red fluorescence (560~nm exc., 635~nm em.) is measured for the WT (dashed lines) and the \textit{rpsB-mCherry} (solid lines) strains.
Fluorescence levels of the modified strains are far above the autofluorescences measured on the WT strain.
\textit{(C)}~Ratio of fluorescence over absorbance, a proxy for the fluorescence concentration in the cells.
Autofluorescence background was corrected by removing the fluorescence measured on the WT.
We observe a strange increase in fluorescence concentration for the \textit{rpsB-mCherry} strain at stationary phase, something that does not occur on the \textit{rpsB-gfp} strain.
At each measurement cycle, the following procedure was applied: shaking (Orbital 6mm) for 30~s, shaking (Linear 6mm) for 30~s, waiting for 5~s.
}
\label{fig:suppinfo_caract}
\end{figure}

In order to explain these strange dynamics, we set-up an experiment to evaluate the degradation and maturation rates of GFP and mCherry in our strains.
The procedure was to instantly stop all protein synthesis in mid-exponential phase using an antibiotic, then to observe the evolution of the fluorescence in a condition with no synthesis of new reporter proteins.
The estimation was performed using the following model:
\begin{eqnarray}
\frac{dP_m}{dt} &=& K \cdot P - (\mu+D) \cdot P_m, \\
\frac{dP}{dt} &=& f(t) - (\mu+D+K) P,
\end{eqnarray}
where $P_m$ is the concentration of mature (fluorescent) proteins, $P$ is the concentration of immature (non-fluorescent) proteins, $K$ is the maturation rate, $D$ is the degradation rate (assumed identical for mature and immature proteins), $f(t)$ the protein synthesis rate, and $\mu$ the growth rate.
By stopping all protein synthesis in the bacteria, we obtain $f(t) = \mu = 0$, hence the model becomes:
\begin{eqnarray}
\frac{dP_m}{dt} &=& K \cdot P - D \cdot P_m, \\
\frac{dP}{dt} &=& - (D+K) P.
\end{eqnarray}
This system of differential equations can be analytically solved into:
\[
P_m(t) = \left(P_m(0) + P(0) \left( 1 - e^{-Kt} \right) \right) e^{-Dt}
\]
with $P_m(0)$ and $P(0)$ the initial concentrations of mature proteins and immature proteins, respectively.
By dividing by $P_m(0)$ and taking the log of both sides, we can rewrite it as
\begin{equation}
\log \frac{P_m(t)}{P_m(0)} = - Dt + \log \left(1 + \frac{P(0)}{P_m(0)} \left( 1 - e^{-Kt} \right) \right).
\label{eq:matur_deg}
\end{equation}
Since growth is stopped, the bacterial volume is constant and then $\frac{P_m(t)}{P_m(0)}$ can be easily measured by directly taking the fluorescence normalized by its initial value, while the maturation rate $K$, the degradation rate $D$, and the initial fraction of immature proteins $\frac{P(0)}{P_m(0)}$ are free parameters that can be fitted on fluorescence time series.
The curves used for the estimation are represented in Fig.~\ref{fig:suppinfo_matur} and the results are reported in Tab.~\ref{tab:suppinfo_matur}.
As expected, both reporter proteins are stable and exhibit a long half-life in exponential phase (>24~h).
The maturation rate of GFP was too fast to obtain numerical value.
However, the maturation rate of mCherry is rather slow, with a half-maturation time in the order of 30~min (time needed to mature 50\% of a pool of mCherry).
Even though degradation and maturation rate can be slightly different in stationary phase, the estimated value are not sufficient to explain the rapid increase in fluorescence observed in Fig.~\ref{fig:suppinfo_caract}.

\begin{figure}[tb]
\centering
\includegraphics[scale=1]{./Fig/suppinfo_matur}
\caption{
\textbf{Estimation of the maturation and degradation rates of the reporter proteins in the \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains.}
The strains were grown in the same condition as in Fig.~\ref{fig:suppinfo_caract}, except that a massive dose of Chloramphenicol was added in mid-exponential phase ($t=0$, vertical dashed grey line).
The fluorescence was normalized by the value at $t=0$ and the degradation and maturation rates were estimated by using the model in Eq.~\ref{eq:matur_deg}.
Parameters of the best fit (solid black line) are reported in Tab.~\ref{tab:suppinfo_matur}.
Data are the result of the aggregation of 9 and 8 independent growth curves for \textit{rpsB-gfp} and \textit{rpsB-mCherry}, respectively.
}
\label{fig:suppinfo_matur}
\end{figure}

@@

The fluorescence of GFP and mCherry is known to be affected by intracellular physiological changes like pH or pO\textsubscript{2}~\cite{doherty_stage-specific_2010}.
In particular, mCherry has been reported to be extremely sensible to the presence of oxygen for its maturation~\cite{doherty_stage-specific_2010}.
We however reproduced the following results in conditions were the oxygen is not limiting.
In particular, the abrupt increase of mCherry fluorescence is conserved when the stationary phase is attained at low bacterial density, in a larger volume (20-mL flask), and on several type of carbon sources (acetate, xylose, glycerol, maltose).
We have so far no explanation for this strange dynamic, and therefore concentrated our effort on the \textit{rpsB-gfp} strain in the microfluidic experiments.

@@

We estimated the noise in the RFU/pixel/cell measurements by using the data points just before the upshift (Fig.~\ref{fig:noise_poi}).
Since they have been growing for 20~h on acetate in the microfluidic device, bacteria are assumed in balanced growth in this region of $\sim$2~hours (roughly 23-24 points depending on the well of interest).
For this reason, ribosome concentration, and so fluorescence concentration, are expected constant.
Fig.~\ref{fig:noise_poi} shows the distribution of points in this region, that can be approximated by a Gaussian distribution, for a particular cell.

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/noise_poi}
\caption{
\textbf{Points of interest for the noise estimation.}\newline
\textit{(Top graph)} RFU/pixel/cell data points for the cell at the bottom of the well labeled X3Y2,W2.
Noise estimation was performed on the points just before the upshift (red points), were the bacteria is assumed to grow at steady state.
\textit{(Bottom graph)} Cumulative density function (CDF, in red) and probability density function (PDF, in gray) of the points highlighted in red on the top graph.
They are visually compared with CDF (solid green line) and PDF (dashed green line) of a Gaussian fit with mean 548.88 and standard deviation 17.316.
}
\label{fig:noise_poi}
\end{figure}

Interestingly, as can be seen in Fig.~\ref{fig:noise_distrib_mean_std}, the mean and standard deviation of the distribution vary from cell to cell in a slightly correlated manner (Pearson R\textsuperscript{2}: $0.3213$, p-value: $4.923\cdot10^{-5}$).
We did not investigate if this heterogeneity between cells is the result of true biological variations or a bias due to the microfluidic device.
By looking at the points of Fig.~\ref{fig:noise_distrib_mean_std}, there are however no strong correlation of the mean and standard deviation with the acquisition field (XY), or the position of the well on the image (W).

@@

Noise characteristics were computed by normalizing and aggregating data for the 45 normal cells presented in Fig.~\ref{fig:noise_distrib_mean_std}.
Two scenarii were considered.
First, the noise model can be considered additive, which means we have the measurement model presented in Section~\ref{sec:estim_gr_ra}:
\begin{equation}
F(t_k) = \gamma r (t_k) + \eta_k,\\
\end{equation}
were $F(t_k)$ is the fluorescence measured at the time step $t_k$, $r (t_k)$ is the true ribosomal concentration at this time step, $\gamma$ is an unknown factor, and $\eta_k$ is the measurement noise.
In this scenario, for a time independent ribosome concentration $r$ (\textit{i.e.} at steady state), the ribosome concentration is equal to its mean on the interval of interest (noted $m_k(\cdot)$).
Given that
\begin{equation}
m_k (F) = m_k (\gamma r) + m_k(\eta_k) = \gamma m_k (r),\\
\end{equation}
the noise residues are given by
\begin{equation}
\eta_k = F(t_k) - m_k (F). \label{eq:noise_add}
\end{equation}
In other words, we can normalize the data points of each cell by removing their mean in order to aggregate the noise residues and get an estimation of the noise level (Fig.~\ref{fig:noise_estim_add_mul}, top panel).

Alternatively, the noise model can be considered multiplicative.
The noise model is than different than presented in Section~\ref{sec:estim_gr_ra}, and can be re-written:
\begin{equation}
F(t_k) = \gamma r (t_k) \cdot (1 + \lambda_k),
\end{equation}
were $\gamma r (t_k) \cdot \lambda_k$ is the measurement noise, proportional to the value measured.
In that scenario, with the same notations as above, the noise $\lambda_k$ is given by
\begin{equation}
\lambda_k = \frac{F(t_k)}{m_k (F)} - 1, \label{eq:noise_mul}
\end{equation}
which means we can normalize data points from different cells by dividing by the mean and removing 1 (Fig.~\ref{fig:noise_estim_add_mul}, bottom panel).

Both models were considered and are represented in Fig.~\ref{fig:noise_estim_add_mul}.
Despite the correlation identified in Fig.~\ref{fig:noise_distrib_mean_std}, the means of the residues in both models are very close to zero, and the distributions are well approximated by a Gaussian distribution.
For this reason, we decided in the main analysis to stick with an additive noise, the main reason being that the implementation used for the Kalman smoothing procedure are faster and more stable with an additive noise model.
We assumed an additive white Gaussian noise with mean 0 and standard deviation 28.29 for the simulation of the synthetic data and the signal reconstruction through the Kalman smoothing procedure.
As discussed in Section~\ref{sec:chap3_discussion}, the only way to decisively choose a noise model over the other would have been to obtain two different steady-state growths for each cell.
Unfortunately, the presented time series were not long enough to obtain a new steady state on glucose, something that will be corrected in future instances of the experiment.

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/noise_estim_add_mul}
\caption{
\textbf{Distribution of the noise residues after normalization and aggregation for the 45 normal cells.}
Two noise models were considered for the normalization: either the temporal mean $m_k (\cdot)$ was removed from the measurements independently for each cell (additive noise model in Eq.~\ref{eq:noise_add}), or $1$ was removed from the ratio of the measurements with the temporal mean $m_k (\cdot)$ (multiplicative noise model in Eq.~\ref{eq:noise_mul}).
For the additive noise model, the fitted Gaussian distribution as a mean of $1.256\cdot 10^{-15}$ and a standard deviation of 28.29.
For the multiplicative noise model, the fitted Gaussian distribution as a mean of $2.666\cdot 10^{-18}$ and a standard deviation of 0.05383.
}
\label{fig:noise_estim_add_mul}
\end{figure}

\clearpage
\subsection{S8 Text -- Examples of complete analysis for 5 normal cells}
\manuallabel{S8_Text}{S8~Text}

In order to get a more visual idea of the results of the analysis, we display below the complete signal reconstructions for 5 cells.
In Figs~\ref{fig:cell1}-\ref{fig:cell5}, the image on the left is the last image analyzed for the corresponding well.
The bacteria either at the top or the bottom (depending on the orientation in the device) was manually segmented by selecting two pixels at the poles on the fluorescence images (red cross).
A 6-pixel-wide rectangular mask was computed for each image, resulting in the masked image on the right.
In each of the graphs, the black and green points represent data points, while the red solid lines are the results of the signal reconstruction \textit{via} the Kalman smoothing procedure.
The parameters of the Kalman smoothing are described in Material and Methods~\ref{sec:meth_kalman} and Figs~\ref{fig:growth_rate_estimation} and \ref{fig:gene_activity}.
Specific comments on each cell are given in the caption of Figs~\ref{fig:cell1}-\ref{fig:cell5}.

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/cell1}
\caption{\small
\textbf{Complete analysis for the cell located at the bottom of the well labelled "X1Y1, W11".}
(Figure description available in the introduction of \ref{S8_Text}.)\newline
Growth rate and resource allocation are particularly unstable at the beginning of the experiment, but seems to have stabilized before the upshift.
Oscillations in the RFU/pixel/cell signal are clearly visible, and result in oscillations in the resource allocation signal reconstruction.
}
\label{fig:cell1}
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/cell4}
\caption{\small
\textbf{Complete analysis for the cell located at the bottom of the well labelled "X2Y2, W3".}
(Figure description available in the introduction of \ref{S8_Text}.)\newline
The reconstruction on this cell is a good instance of what was expected: the growth rate is stable on acetate, than quickly increases after the upshift on glucose.
The resource allocation is stable on acetate (indicating a steady state) than starts to oscillate after the upshift on glucose, even though the smoothing factor seems a little too high for this particular cell (some oscillations are poorly predicted after the upshift, in particular in the interval [200,400]~min).
}
\label{fig:cell4}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/cell5}
\caption{\small
\textbf{Complete analysis for the cell located at the bottom of the well labelled "X1Y2, W6".}
(Figure description available in the introduction of \ref{S8_Text}.)\newline
The results on this cell are particularly terrible: we do not seem to get a good steady state on acetate since the reconstructed resource allocation appears either negative or oscillating before the upshift.
It is thus impossible to conclude if the oscillations observed after the upshift are real or due to overfitting.
This cell highlights why robust steady states before and after the transition are important to conclude on the signal reconstruction.
}
\label{fig:cell5}
\end{figure}

\clearpage
\subsection{S2 Figure -- Cell categories identified in the microscopy analysis}
\manuallabel{S2_Figure}{S2~Fig}

>>>>

The following pages contain the full annotated sequences of the region downstream of \textit{rpsB} after the insertion of the \textit{gfp\_mut2-cassette} and the \textit{mCherry-cassette}, respectively (third line in Fig.~\ref{fig:Cloning}).
The positions where the primers anneal to the template are indicated using the same color code as in the description of the primers above.
The primers reported in the publication of Bakshi \textit{et al.}~\cite{bakshi_superresolution_2012} are indicated for information.

@@

After the strains were constructed using the protocol described in Material and Methods~\ref{sec:methods_strain}, the downstream region of the \textit{rpsB} gene was amplified and sequenced.
Alignments of the results confirmed the expected sequences, except for a few single-base mutations.
The raw results of the sequencing are reproduced below.

@@

In Fig.~\ref{fig:suppinfo_caract}, we show growth curves obtained on M9 supplemented with 0.2\% glucose.
As for the construction of Bakshi \textit{et al.}~\cite{bakshi_superresolution_2012}, our strains possess a wild-type growth rate (Fig.~\ref{fig:suppinfo_caract}A), indicating that the gfp-tagging or mCherry-tagging of the ribosomal S2 subunit does not impede the functioning of the ribosome.
It is known that \textit{E.~coli} cultures exhibit a high autofluorescence overlapping the GFP spectrum~\cite{mihalcescu_green_2015}, but the fluorescence of the \textit{rpsB-gfp} strain is at least 4-time higher than the \textit{wild-type} (Fig.~\ref{fig:suppinfo_caract}B).
Contrary to green autofluorescence, red autofluorescence of \textit{E.~coli} cultures is very low (Fig.~\ref{fig:suppinfo_caract}B) so the corrected signal level for mCherry is even stronger.
Overall, both strains possess enough fluorescence to allow quantification at the population level, all the more so at the cell level because most of the autofluorescence is concentrated in the medium outside the cell~\cite{mihalcescu_green_2015}.

In Fig.~\ref{fig:suppinfo_caract}C, we display the ratio between the (corrected) fluorescence of the strain and the absorbance ("corrected" meaning that we substracted the autofluorescence of the wild-type strain).
As a first approximation, this ratio can be used as a proxy for the fluorescence concentration in the cells, therefore for the ribosome concentration.
The instability at the beginning of the experiment (before 400~min) is characteristic of the difficulty of obtaining robust estimations when the absorbance of the culture is low~\cite{zulkower_robust_2015}.
The RFU/Abs ratio is stable in the interval [400,600]~min for both \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains, indicative of exponential steady-state growth.
Note that this short time interval (a couple of generations) of readability before approaching the stationary phase is one of the main problems preventing the use of batch conditions for obtaining robust data about growth transitions.
Indeed, most of our attempts to perform robust growth transitions in this region were unsuccessful.

Results for \textit{rpsB-gfp} and \textit{rpsB-mCherry} start to differ when glucose is exhausted and the cells enter stationary phase.
As would be expected, the fluorescence of the \textit{rpsB-gfp} strain starts to slowly decrease as GFP proteins are degraded or photobleached.
On the contrary, the fluorescence of the \textit{rpsB-mCherry} strain quickly triples when entering stationary phase, at a rate that is much higher than the physical limit imposed by the protein synthesis rate~\cite{scott_interdependence_2010}.

\begin{figure}[p]
\centering
\includegraphics[scale=0.985]{./Fig/suppinfo_caract}
\caption{\small
\textbf{Growth curves in M9~0.2\% glucose for the \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains.}
Growth and measurements were performed at 37$^\circ$C on a 96-well microplate in a Tecan infinite 200 pro.
We monitored 4 wells for each of the WT, \textit{rpsB-gfp}, and \textit{rpsB-mCherry} strains.
Curves were time shifted to correct for variability in the inoculation process (-50 min for WT, and -20 min for \textit{rpsB-mCherry}).
\textit{(A)}~Absorbance at 600~nm, corrected for background by subtracting the absorbance of the M9 medium.
Curve superposition indicates similar growth rates between the WT and the modified strains.
\textit{(B)}~Fluorescence measured in each well in Relative Fluorescence Units (RFU).
Green fluorescence (485~nm excitation, 535~nm emission) is measured for the WT (dashed lines) and the \textit{rpsB-gfp} (solid lines) strains.
Red fluorescence (560~nm exc., 635~nm em.) is measured for the WT (dashed lines) and the \textit{rpsB-mCherry} (solid lines) strains.
Fluorescence levels of the modified strains are far above the autofluorescences measured on the WT strain.
\textit{(C)}~Ratio of fluorescence over absorbance (proxy for the fluorescence concentration in the cells).
Autofluorescence background was corrected by removing the fluorescence measured on the WT.
We observed a strange increase in fluorescence concentration for the \textit{rpsB-mCherry} strain after entry into stationary phase, something that does not occur on the \textit{rpsB-gfp} strain.
At each measurement cycle, the following procedure was applied: shaking (Orbital 6mm) for 30~s, shaking (Linear 6mm) for 30~s, waiting for 5~s.
}
\label{fig:suppinfo_caract}
\end{figure}

In order to explain these strange dynamics, we set-up an experiment to evaluate the degradation and maturation rates of GFP and mCherry in our strains.
The procedure was to instantly stop all protein synthesis in mid-exponential phase using an antibiotic, then to observe the evolution of the fluorescence in a condition with no synthesis of new reporter proteins.
The estimation was performed using the following model:
\begin{eqnarray}
\frac{dP_m}{dt} &=& K \cdot P - (\mu+D) \cdot P_m, \\
\frac{dP}{dt} &=& f(t) - (\mu+D+K) P,
\end{eqnarray}
where $P_m$ is the concentration of mature (fluorescent) proteins, $P$ is the concentration of immature (non-fluorescent) proteins, $K$ is the maturation rate, $D$ is the degradation rate (assumed identical for mature and immature proteins), $f(t)$ the protein synthesis rate, and $\mu$ the growth rate.
By stopping all protein synthesis in the bacteria, we obtain $f(t) = \mu = 0$, hence the model becomes:
\begin{eqnarray}
\frac{dP_m}{dt} &=& K \cdot P - D \cdot P_m, \\
\frac{dP}{dt} &=& - (D+K) P.
\end{eqnarray}
This system of differential equations can be analytically solved to yield:
\[
P_m(t) = \left(P_m(0) + P(0) \left( 1 - e^{-Kt} \right) \right) e^{-Dt}
\]
with $P_m(0)$ and $P(0)$ the initial concentrations of mature proteins and immature proteins, respectively.
By dividing by $P_m(0)$ and taking the log of both sides, we can rewrite it as
\begin{equation}
\log \frac{P_m(t)}{P_m(0)} = - Dt + \log \left(1 + \frac{P(0)}{P_m(0)} \left( 1 - e^{-Kt} \right) \right).
\label{eq:matur_deg}
\end{equation}
Since growth is stopped, the bacterial volume is constant and $\frac{P_m(t)}{P_m(0)}$ can now be easily measured by directly taking the fluorescence normalized by its initial value, while the maturation rate $K$, the degradation rate $D$, and the initial fraction of immature proteins $\frac{P(0)}{P_m(0)}$ are free parameters that can be fitted on fluorescence time series.
The curves used for the estimation are represented in Fig.~\ref{fig:suppinfo_matur} and the results are reported in Tab.~\ref{tab:suppinfo_matur}.
As expected, both reporter proteins are stable and exhibit a long half-life in exponential phase (>24~h).
The maturation rate of GFP was too fast for obtaining a numerical value.
However, the maturation rate of mCherry is rather slow, with a half-maturation time on the order of 30~min (time needed to mature 50\% of a pool of mCherry).
Even though degradation and maturation rate can be slightly different in stationary phase, the estimated value are not sufficient to explain the rapid increase in fluorescence observed in Fig.~\ref{fig:suppinfo_caract}.

\begin{figure}[tb]
\centering
\includegraphics[scale=1]{./Fig/suppinfo_matur}
\caption{
\textbf{Estimation of the maturation and degradation rates of the reporter proteins in the \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains.}
The strains were grown in the same condition as in Fig.~\ref{fig:suppinfo_caract}, except that a high concentration of Chloramphenicol was added in mid-exponential phase ($t=0$, vertical dashed grey line).
The fluorescence was normalized by the value at $t=0$ and the degradation and maturation rates were estimated by using the model in Eq.~\ref{eq:matur_deg}.
Parameters of the best fit (solid black line) are reported in Tab.~\ref{tab:suppinfo_matur}.
Data are the result of the aggregation of 9 and 8 independent growth curves for \textit{rpsB-gfp} and \textit{rpsB-mCherry}, respectively.
}
\label{fig:suppinfo_matur}
\end{figure}

@@

The fluorescence of GFP and mCherry is known to be affected by intracellular physiological changes like pH or pO\textsubscript{2}~\cite{doherty_stage-specific_2010}.
In particular, mCherry has been reported to be extremely sensible to the presence of oxygen for its maturation~\cite{doherty_stage-specific_2010}.
However, we reproduced the following results in conditions were the oxygen is not limiting.
In particular, the abrupt increase of mCherry fluorescence is conserved when the stationary phase is attained at low bacterial density, in a larger volume (20-mL flask), and on several type of carbon sources (acetate, xylose, glycerol, maltose).
We have so far no explanation for these strange dynamics, and we therefore concentrated our efforts on the \textit{rpsB-gfp} strain in the microfluidic experiments.

@@

We estimated the noise in the RFU/pixel/cell measurements by using the data points just before the upshift (Fig.~\ref{fig:noise_poi}).
Since the cells have been growing for 20~h on acetate in the microfluidic device, bacteria are assumed to be in balanced growth in this region of $\sim$2~hours (roughly 23-24 points depending on the well of interest).
For this reason, ribosome concentration, and therefore the fluorescence concentration, are expected to be constant.
Fig.~\ref{fig:noise_poi} shows the distribution of points in this region, which can be approximated by a Gaussian distribution, for a particular cell.

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/noise_poi}
\caption{
\textbf{Points of interest for the noise estimation.}\newline
\textit{(Top graph)} RFU/pixel/cell data points for the cell at the bottom of the well labeled X3Y2,W2.
Noise estimation was performed on the points just before the upshift (red points), were the bacteria are assumed to grow at steady state.
\textit{(Bottom graph)} Cumulative density function (CDF, in red) and probability density function (PDF, in gray) of the points highlighted in red on the top graph.
They are visually compared with CDF (solid green line) and PDF (dashed green line) of a Gaussian fit with mean 548.88 and standard deviation 17.316.
}
\label{fig:noise_poi}
\end{figure}

Interestingly, as can be seen in Fig.~\ref{fig:noise_distrib_mean_std}, the mean and standard deviation of the distribution vary from cell to cell in a slightly correlated manner (Pearson R\textsuperscript{2}: $0.3213$, p-value: $4.923\cdot10^{-5}$).
We did not investigate if this heterogeneity between cells is the result of true biological variations or a bias due to the microfluidic device.
By looking at the points of Fig.~\ref{fig:noise_distrib_mean_std}, there are however no strong correlations of the mean and standard deviation with the acquisition field (XY), or the position of the well on the image (W).

@@

Noise characteristics were computed by normalizing and aggregating data for the 45 normal cells presented in Fig.~\ref{fig:noise_distrib_mean_std}.
Two scenarios were considered.
First, the noise is additive, which means that we have the measurement model presented in Section~\ref{sec:estim_gr_ra}:
\begin{equation}
F(t_k) = \gamma r (t_k) + \eta_k,\\
\end{equation}
were $F(t_k)$ is the fluorescence measured at the time step $t_k$, $r (t_k)$ is the true ribosomal concentration at this time step, $\gamma$ is an unknown factor, and $\eta_k$ is the measurement noise.
In this scenario, for a time independent ribosome concentration $r$ (\textit{i.e.} at steady state), the ribosome concentration is equal to its mean on the interval of interest (noted $m_k(\cdot)$).
Given that
\begin{equation}
m_k (F) = m_k (\gamma r) + m_k(\eta_k) = \gamma m_k (r),\\
\end{equation}
the noise residues are given by
\begin{equation}
\eta_k = F(t_k) - m_k (F). \label{eq:noise_add}
\end{equation}
In other words, we can normalize the data points of each cell by removing their mean in order to aggregate the noise residues and thus obtain an estimation of the noise level (Fig.~\ref{fig:noise_estim_add_mul}, top panel).

Alternatively, the noise could be multiplicative.
The noise model is than different of the one presented in Section~\ref{sec:estim_gr_ra}, and can be re-written:
\begin{equation}
F(t_k) = \gamma r (t_k) \cdot (1 + \lambda_k),
\end{equation}
were $\gamma r (t_k) \cdot \lambda_k$ is the measurement noise, proportional to the value measured.
In that scenario, with the same notations as above, the noise $\lambda_k$ is given by
\begin{equation}
\lambda_k = \frac{F(t_k)}{m_k (F)} - 1, \label{eq:noise_mul}
\end{equation}
which means we can normalize data points from different cells by dividing by the mean and removing 1 (Fig.~\ref{fig:noise_estim_add_mul}, bottom panel).

Both models were considered and are represented in Fig.~\ref{fig:noise_estim_add_mul}.
Despite the correlation identified in Fig.~\ref{fig:noise_distrib_mean_std}, the means of the residues in both models are very close to zero, and the distributions are well approximated by a Gaussian distribution.
For this reason, we decided in the main analysis to stick with an additive noise, the main reason being that the implementation used for the Kalman smoothing procedure are faster and more stable with an additive noise model.
We assumed an additive white Gaussian noise with mean 0 and standard deviation of 28.29 for the simulation of the synthetic data and the signal reconstruction through the Kalman smoothing procedure.
As discussed in Section~\ref{sec:chap3_discussion}, the only way to definitively choose a noise model over the other would have been to obtain two different steady-state growths for each cell.
Unfortunately, the presented time series were not long enough to obtain a new steady state on glucose, something that will be corrected in future experiments.

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/noise_estim_add_mul}
\caption{
\textbf{Distribution of the noise residues after normalization and aggregation for the 45 normal cells.}
Two noise models were considered for the normalization: either the temporal mean $m_k (\cdot)$ was removed from the measurements independently for each cell (additive noise model in Eq.~\ref{eq:noise_add}), or $1$ was removed from the ratio of the measurements with the temporal mean $m_k (\cdot)$ (multiplicative noise model in Eq.~\ref{eq:noise_mul}).
For the additive noise model, the fitted Gaussian distribution has a mean of $1.256\cdot 10^{-15}$ and a standard deviation of 28.29.
For the multiplicative noise model, the fitted Gaussian distribution has a mean of $2.666\cdot 10^{-18}$ and a standard deviation of 0.05383.
}
\label{fig:noise_estim_add_mul}
\end{figure}

\clearpage
\subsection{S8 Text -- Examples of complete analysis for 5 cells}
\manuallabel{S8_Text}{S8~Text}

In order to obtain a more visual idea of the results of the analysis, we display below the complete signal reconstructions for 5 cells.
In Figs~\ref{fig:cell1}-\ref{fig:cell5}, the image on the left is the last image analyzed for the corresponding well.
The cell either at the top or the bottom (depending on the orientation in the device) was manually segmented by selecting two pixels at the poles on the fluorescence images (red cross).
A 6-pixel-wide rectangular mask was computed for each image, resulting in the masked image on the right.
In each of the graphs, the black and green points represent data points, while the red solid lines are the results of the signal reconstruction \textit{via} the Kalman smoothing procedure.
The parameters of the Kalman smoothing are described in Material and Methods~\ref{sec:meth_kalman} and Figs~\ref{fig:growth_rate_estimation} and \ref{fig:gene_activity}.
Specific comments on each cell are given in the caption of Figs~\ref{fig:cell1}-\ref{fig:cell5}.

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/cell1}
\caption{\small
\textbf{Complete analysis for the cell located at the bottom of the well labelled "X1Y1, W11".}
(Figure description available in the introduction of \ref{S8_Text}.)\newline
Growth rate and resource allocation are particularly unstable at the beginning of the experiment, but seem to stabilize before the upshift.
Oscillations in the RFU/pixel/cell signal are clearly visible and result in oscillations in the resource allocation signal reconstruction.
}
\label{fig:cell1}
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/cell4}
\caption{\small
\textbf{Complete analysis for the cell located at the bottom of the well labelled "X2Y2, W3".}
(Figure description available in the introduction of \ref{S8_Text}.)\newline
The reconstruction on this cell is a good example of what was expected: the growth rate is stable on acetate, than quickly increases after the upshift on glucose.
The resource allocation is stable on acetate (indicating a steady state) than starts to oscillate after the upshift on glucose, even though the smoothing factor seems a little too high for this particular cell (some oscillations are poorly predicted after the upshift, in particular in the interval [200,400]~min).
}
\label{fig:cell4}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/cell5}
\caption{\small
\textbf{Complete analysis for the cell located at the bottom of the well labelled "X1Y2, W6".}
(Figure description available in the introduction of \ref{S8_Text}.)\newline
The results on this cell are particularly questionable: we do not seem to obtain a good steady state on acetate since the reconstructed resource allocation appears either negative or oscillating before the upshift.
It is thus impossible to conclude if the oscillations observed after the upshift are real or due to overfitting.
This cell highlights why robust steady states before and after the transition are important for a reliable signal reconstruction.
}
\label{fig:cell5}
\end{figure}

\clearpage
\subsection{S2 Figure -- Cell categories identified in the analysis}
\manuallabel{S2_Figure}{S2~Fig}

\begin{figure}[!h]
\centering
\includegraphics[scale=1]{./Fig/subcat_cells}
\caption{
\textbf{Cell categories identified in the microscopy analysis.}
As stated in Section~\ref{sec:estim_gr_ra}, the reconstruction of the growth rate for the 68 available cells motivated their classification in three categories: dying cells stop growing (lowest graphs) , pausing cells present growth rate oscillations (middle graph), and normal cells that do not present any of the above (highest graph).}
\label{fig:subcat_cells}
\end{figure}
